*dfoliatR*: An R package for detection and analysis of insect defoliation signals in tree rings

### 2020-05-11 paper_elsevier/paper_elsevier.Rmd
edits before send to Jodi

---
title: "*dfoliatR*: An R package for detection and analysis of insect defoliation signals in tree rings"
author:
  - name: Christopher H. Guiterman
    email: chguiterman@arizona.edu
    affiliation: a,b
    footnote: 1
  - name: Ann M. Lynch
    email: ann.m.lynch@usda.gov
    affiliation: a,c
    # footnote: [2, 3]
  - name: Jodi N. Axelson
    email: jodi.axelson@berkeley.edu
    affiliation: c
    # footnote: 4
address:
  - code: a
    address: "Laboratory of Tree-Ring Research, University of Arizona, 1215 E Lowell St. Box 210045, Tucson, AZ, 85721"
  - code: b
    address: "Three Pines Forest Research, LLC, PO Box 225, Etna, NH, 03750"
  - code: c
    address: "U.S. Forest Service, Rocky Mountain Research Station, 1215 E Lowell St. Box 210045, Tucson, AZ, 85721"
  - code: d
    address: "Dept of Environmental Science, Policy & Management, University of California, Berkeley, Berkeley, CA 94720"
footnote:
  - code: 1
    text: "Corresponding Author"
#   - code: 2
#     text: "Laboratory of Tree-Ring Research, University of Arizona, 1215 E Lowell St. Box 210045, Tucson, AZ, 85721"
#   - code: 3
#     text: "U.S. Forest Service, Rocky Mountain Research Station, 1215 E Lowell St. Box 210045, Tucson, AZ, 85721"
#   - code: 4
#     text: "Dept of Environmental Science, Policy & Management, University of California, Berkeley, Berkeley, CA 94720"
abstract: |
 We present a new R package to provide dendroecologists with tools to infer, quantify, analyze, and visualize growth suppression events in tree rings. `dfoliatR` is based on the OUTBREAK program and builds on existing resources in the `R` and `dplR` computing environments. It is designed wo aid research in the ecology of insect defoliation events and to reconstruct defoliator outbreak chronologies, but can be applied to other studies where host--non-host comparisons are useful. `It`dfoliatR` performs an indexing procedure to remove climatic signals in the host-tree series that are represented in the non-host chronologies, or other annually-resolved climate series. It then infers defoliation events in individual trees based on user-specified thresholds.  Site-level analyses identify outbreak events that synchronously affect user-defined numbers or proportions of involved host trees. Functions are provided for summary statistics and graphics of tree- and site-level series.  `dfoliatR` expands on OUTBREAK to provide greater control of suppression thresholds, additional output tables, and high-quality graphics. We provide two example data sets and script to enable users to gain familiarity with the package and its capabilities. The source code is available in the Comprehensive R Archive Network (CRAN) and on GitHub.
 \hfill\break
keywords: |
  Dendroecology, spruce budworm, *Choristoneura*, pandora moth, *Coloradia pandora Blake*, larch-bud-moth
  \newpage
journal: "Dendrochronologia"
date: "`r Sys.Date()`"
bibliography: References.bib
linenumbers: true
#numbersections: true
# csl: ecology.csl
# csl: elsevier-harvard.csl
# csl: elsevier-vancouver.csl
# output: rticles::elsevier_article
output:
  bookdown::pdf_book:
      base_format: rticles::elsevier_article
#   bookdown::word_document2:
#     reference_docx: "../common/word-styles-reference-01.docx"
#     toc: false
#     toc_depth: 2
#     fig_caption: yes
#     df_print: kable
# always_allow_html: true
layout: review
preamble: |
  \usepackage{amsmath}
  \usepackage{float}
  \usepackage{booktabs}
---

@@

`dfoliatR` adds to a growing suite of dendrochronology packages in the `R` computing environment [@RCore]. Stemming from the `dplR` library [@Bunn2008] that enables `R` users to read and write an array of tree-ring data formats, standardize ring width series, build and evaluate chronologies, and perform quality control (to name a few), one can now also measure ring widths from scanned images of prepared samples [@Shi2019; @Lara2015], conduct and check crossdating [@Bunn2010], analyze sub-annual anotomical features [@Campelo2016; @Rathgeber2011], and perform many analytical tests [@Jevsenak2018; @Zang2015]. Tools for assessing stand dynamics and disturbance analyses are under rapid development, with new packages for assessing growth and release events [`TRADER`: @Altman2014], metrics of growth resilience [`pointRes`: @VanderMaaten-Theunissen2015], and fire history [`burnr`: @Malevich2018]. The key objective of `dfoliatR` is to provide tools to identify and analyze insect defoliation and outbreak events by building on the methods employed by OUTBREAK. It capitalizes on the robust software already available in `R` by using `dplR` data formats for incoming tree-ring series and providing output data formats embodied by the `tidyverse` [@Wickham2019] that include efficient data manipulation [`dplyr`: @Wickham2020dplyr] and graphics [`ggplot2`: @wickham2016ggplot2].

In this paper, we describe the statistical methods employed by `dfoliatR`, compare results to those produced by OUTBREAK, and present an example analysis inlcuding test data sets and script. Users need not have much experience in `R` to replicate the analyses and graphics as presented. The `R` code below is executable in an `R` session once the required libraries are installed and loaded. Support documentation in addition to this paper is provided within the package via standard help menus and on the package website ([https://chguiterman.github.io/dfoliatR/](https://chguiterman.github.io/dfoliatR/)), which includes up-to-date vignettes that describe various routines. Code to generate a preprint of this manuscript, including the `R` scripts and tabular and graphical output is available from [https://github.com/chguiterman/dfoliatR_paper](https://github.com/chguiterman/dfoliatR_paper).

@@

@@

Negative departures in the normalized GSI (NGSI, or GSI converted to z-scores) that surpass user-specified thresholds in duration and magnitude are defined as *defoliation events*. As in OUTBREAK, the lowest NGSI value in the particucar sequence being assessed must reach the magnitude threshold. The default setting is -1.28 (NGSI is in units of standard deviation), which was previously determined to be representative of WSBW effects (Swetnam and Lynch, 1989) and is commonly used for other species. The year with the lowest value is termed the "year of maximum departure" and becomes a central point in time for assessing other thresholds before being included as a defoliation event. If the year of maximum departure is higher than the threshold (i.e. $\textrm{NGSI}_{lowest} > -1.28$), the sequence being assessed is omitted from the event results.

@@

Diverging from OUTBREAK, `dfoliatR` allows users to extend defoliation events on individual trees by bridging between sequential events. In cases where two defoliation events are separated by a single year, bridging will link them into a single event. This option was added to `dfoliatR` during the testing phase of development, when we realized that OUTBREAK deliberately omits sequential, or back-to-back events, even when these events surpass both the magnitude and duration thresholds. Instead, OUTBREAK will select the one sequential event with the lowest negative departure year. In every case we assessed (described below) we felt that the OUTBREAK-omitted defoliation events should have been maintained and recorded. Due in large part to reconstructions using OUTBREAK [see papers cited by @Lynch2012], we now know considerably more about forest defoliator outbreak regimes than we did in the 1980s when OUTBREAK was written.  We think that two or more prolonged events separated by a single year should in some situations be considered a single event. This is particularly relevant to WSBW and spruce budworm (*C. fumiferana*), for which outbreak regimes have been reconstructed for several geographic areas and for which researchers and forest health experts have gained considerable knowledge [@Schmitt1984; @Sanders1985; @Brookes1987, and many later publications].  In many cases, the greatest growth suppression often occurs late in the outbreak due to cumulative effects accrued on a tree's resources [@Swetnam1989; @Swetnam1993]. We urge caution in using the bridging option, however, because it may not be appropriate for all studied insects, such as in situations where impacted stands barely recover from one outbreak before another begins, as with pine processionary caterpillars (*Thaumetopoea pityocampa* (Lepidoptera: Thaumetopoeidae)) [@Carus2004; @Carus2009].

@@

* *Sequential events* (36%) in which OUTBREAK omitted back-to-back events that occur one year prior or one year following an identified event. When this occurs, OUTBREAK selects the one event sequence with the lowest negative index year. On two trees, OUTBREAK omitted two of three sequential events. While inspecting these differences, we added an option to `defoliate_trees()` that would "bridge" between sequential events (that surpass magnitude and duration thresholds) into single, long events. We felt that this was ecologically justified, especially for studies of WSBW, because the insect will occasionally defoliate individual trees for decades and reconstructed chronologies have show outbreaks as long as **xxx (ANN CHECK)**.

@@

**(Bridging paragraph?)**

At the site level, OUTBREAK and `dfoliatR` produce similar time series of percent trees defoliated (Figure \@ref(fig:df-obr-site)), which forms the basis for inferring outbreak occurrence, intensity, and duration. In nearly site comparison, `dfoliatR` included either more events or longer durations of inferred outbreak. These differences arise from the inclusion of tree-level events by `dfoliatR` that were omitted by OUTBREAK. The differences translate to greater numbers of trees in defoliation during certain events, or events recorded by single trees during periods of low sample depth.

@@

The `dfoliatR` package provides dendroecologists with tools to infer, quantify, analyze, and visualize tree-ring growth suppression events and to reconstruct forest insect defoliator outbreak regimes.  It is built on the long-accepted host to nonhost comparison methodology used in the 1980s FORTRAN program OUTBREAK [@Swetnam1985; @Swetnam1989; @Swetnam1993]. Our evaluation of the two programs revelaed that `dfolaitR` excelled in identifying defoliation events on single trees, providing researchers with more consistent and biologically-justifiable results. Other key benefits are that `dfoliatR` provides easier control of suppression thresholds, additional output tables, high-quality and customizable graphics, and operates in the open-source `R` environment that is stable across computing platforms and under active development and maintence by a large and growing community.

@@

<div id="refs"></div>
\newpage

# Tables & Figures {-}

```{r fig-defol, echo=FALSE, fig.cap="Default dfoliatR graphics for individual trees, produced by the 'plot_defol()' function. Dotted lines represent the series length for each tree and colored segments show periods of defoliation. Users can define the Cut-off values to determine severe-moderate-minor defoliation intensities, and use standard ggplot2 graphics parameters to adjust styles and themes."}
include_graphics(here("paper_elsevier", "Output", "tree-plot-default.pdf"))
```

```{r fig-obr, echo=FALSE, fig.cap="Default graphic to show outbreak events at the DMJ site. Top panel shows the mean GSI through time, with inferred outbreak events filled. The middle panel shows the percent of trees defoliated, and the bottom panel shows the number of trees through time."}
include_graphics(here("paper_elsevier", "Output", "site-plot.pdf"))
```

``` {r df-obr-site, echo=FALSE, fig.cap="Comparison of reconstructed western spruce budworm outbreaks computed by dfoliatR and OUTBREAK. Input parameters were identical between programs."}
include_graphics(here("paper_elsevier", "Output", "plot-sites-df-obr.pdf"))
```

>>>>

---
title: "*dfoliatR*: An R package for detection and analysis of insect defoliation signals in tree rings"
author:
  - name: Christopher H. Guiterman
    email: chguiterman@arizona.edu
    affiliation: a,b
    footnote: 1
  - name: Ann M. Lynch
    email: ann.m.lynch@usda.gov
    affiliation: a,c
    # footnote: [2, 3]
  - name: Jodi N. Axelson
    email: jodi.axelson@berkeley.edu
    affiliation: c
    # footnote: 4
address:
  - code: a
    address: "Laboratory of Tree-Ring Research, University of Arizona, 1215 E Lowell St. Box 210045, Tucson, AZ, 85721"
  - code: b
    address: "Three Pines Forest Research, LLC, PO Box 225, Etna, NH, 03750"
  - code: c
    address: "U.S. Forest Service, Rocky Mountain Research Station, 1215 E Lowell St. Box 210045, Tucson, AZ, 85721"
  - code: d
    address: "Dept of Environmental Science, Policy & Management, University of California, Berkeley, Berkeley, CA 94720"
footnote:
  - code: 1
    text: "Corresponding Author"
#   - code: 2
#     text: "Laboratory of Tree-Ring Research, University of Arizona, 1215 E Lowell St. Box 210045, Tucson, AZ, 85721"
#   - code: 3
#     text: "U.S. Forest Service, Rocky Mountain Research Station, 1215 E Lowell St. Box 210045, Tucson, AZ, 85721"
#   - code: 4
#     text: "Dept of Environmental Science, Policy & Management, University of California, Berkeley, Berkeley, CA 94720"
abstract: |
 We present a new R package to provide dendroecologists with tools to infer, quantify, analyze, and visualize growth suppression events in tree rings. `dfoliatR` is based on the OUTBREAK program and builds on existing resources in the `R` and `dplR` computing environments. It is designed to aid research in the ecology of insect defoliation events and to reconstruct defoliator outbreak chronologies, but can be applied to other studies where host--non-host comparisons are useful. `dfoliatR` performs an indexing procedure to remove climatic signals in the host-tree series that are represented in the non-host chronologies, or other annually-resolved climate series. It then infers defoliation events in individual trees based on user-specified thresholds.  Site-level analyses identify outbreak events that synchronously affect user-defined numbers or proportions of involved host trees. Functions are provided for summary statistics and graphics of tree- and site-level series. We evaluated `dfoliatR` against OUTBREAK, using eight datasets including 222 host-trees, and found that `dfoliatR` improves on OUTBREAK with greater user control, identification of defoliation events, computing capacity, and both the statistical summary and graphical outputs. We provide two example data sets and script to enable users to gain familiarity with the package and its capabilities. The source code is available in the Comprehensive R Archive Network (CRAN) and on GitHub.
  <!-- \hfill\break -->
keywords: |
  Dendroecology, spruce budworm, *Choristoneura*, pandora moth, *Coloradia pandora Blake*, larch-bud-moth
  <!-- \newpage -->
# journal: "Dendrochronologia"
# date: "`r Sys.Date()`"
bibliography: References.bib
linenumbers: true
#numbersections: true
# csl: ecology.csl
# csl: elsevier-harvard.csl
# csl: elsevier-vancouver.csl
csl: dendrochronologia.csl
# output: rticles::elsevier_article
# output:
#   bookdown::pdf_book:
#       base_format: rticles::elsevier_article
bookdown::word_document2:
    reference_docx: "../common/word-styles-reference-01.docx"
    toc: false
#     toc_depth: 2
    fig_caption: yes
    df_print: kable
# always_allow_html: true
layout: review
preamble: |
  \usepackage{amsmath}
  \usepackage{float}
  \usepackage{booktabs}
  \makeatletter
  \def\ps@pprintTitle{%
   \let\@oddhead\@empty
   \let\@evenhead\@empty
   \let\@oddfoot\@empty
   \let\@evenfoot\@oddfoot
  }
  \makeatother
---

@@

`dfoliatR` adds to a growing suite of dendrochronology packages in the `R` computing environment [@RCore]. Stemming from the `dplR` library [@Bunn2008] that enables `R` users to read and write an array of tree-ring data formats, standardize ring width series, build and evaluate chronologies, and perform quality control (to name a few), one can now also measure ring widths from scanned images of prepared samples [@Shi2019; @Lara2015], conduct and check crossdating [@Bunn2010], analyze sub-annual anatomical features [@Campelo2016; @Rathgeber2011], and perform many analytical tests [@Jevsenak2018; @Zang2015]. Tools for assessing stand dynamics and disturbance analyses are under rapid development, with new packages for assessing growth and release events [`TRADER`: @Altman2014], metrics of growth resilience [`pointRes`: @VanderMaaten-Theunissen2015], and fire history [`burnr`: @Malevich2018]. The key objective of `dfoliatR` is to provide tools to identify and analyze insect defoliation and outbreak events by building on the methods employed by OUTBREAK. It capitalizes on the robust software already available in `R` by using `dplR` data formats for incoming tree-ring series and providing output data formats embodied by the `tidyverse` [@Wickham2019] that include efficient data manipulation [`dplyr`: @Wickham2020dplyr] and graphics [`ggplot2`: @wickham2016ggplot2].

In this paper, we describe the statistical methods employed by `dfoliatR`, compare results to those produced by OUTBREAK, and present an example analysis including test data sets and script. Users need not have much experience in `R` to replicate the analyses and graphics as presented. The `R` code below is executable in an `R` session once the required libraries are installed and loaded. Support documentation in addition to this paper is provided within the package via standard help menus and on the package website ([https://chguiterman.github.io/dfoliatR/](https://chguiterman.github.io/dfoliatR/)), which includes up-to-date vignettes that describe various routines. Code to generate a preprint of this manuscript, including the `R` scripts and tabular and graphical output is available from [https://github.com/chguiterman/dfoliatR_paper](https://github.com/chguiterman/dfoliatR_paper).

@@

(ref:defolCap) Default `dfoliatR` graphics for individual trees, produced by the `plot_defol()` function. Dotted lines represent the series length for each tree and colored segments show periods of defoliation. Users can define the cut-off values to determine severe--moderate--minor defoliation intensities, and use standard `ggplot2` graphics parameters to adjust styles and themes.

```{r fig-defol, echo=FALSE, fig.cap='(ref:defolCap)'}
include_graphics(here("paper_elsevier", "Output", "tree-plot-default.pdf"))
```

(ref:obrCap) Default graphics showing outbreak events at the DMJ site. Top panel shows the mean GSI through time, with inferred outbreak events filled. The middle panel shows the percent of trees defoliated, and the bottom panel shows the number of trees through time.

```{r fig-obr, echo=FALSE, fig.cap='(ref:obrCap)'}
include_graphics(here("paper_elsevier", "Output", "site-plot.pdf"))
```

@@

Negative departures in the normalized GSI (NGSI, or GSI converted to z-scores) that surpass user-specified thresholds in duration and magnitude are defined as *defoliation events*. As in OUTBREAK, the lowest NGSI value in the particular sequence being assessed must reach the magnitude threshold. The default setting is -1.28 (NGSI is in units of standard deviation), which was previously determined to be representative of WSBW effects (Swetnam and Lynch, 1989) and is commonly used for other species. The year with the lowest value is termed the "year of maximum departure" and becomes a central point in time for assessing other thresholds before being included as a defoliation event. If the year of maximum departure is higher than the threshold (i.e. $\textrm{NGSI}_{lowest} > -1.28$), the sequence being assessed is omitted from the event results.

@@

Diverging from OUTBREAK, `dfoliatR` allows users to extend defoliation events on individual trees by bridging between sequential events (Figure \@ref(fig:fig-bridge)). In cases where two defoliation events are separated by a single year, bridging will link them into a single event. This option was added to `dfoliatR` during the testing phase of development, when we realized that OUTBREAK deliberately omits sequential, or back-to-back events, even when these events surpass both the magnitude and duration thresholds. Instead, OUTBREAK will select the one sequential event with the lowest negative departure year. In every case we assessed (described below) we felt that the OUTBREAK-omitted defoliation events should have been maintained and recorded. Due in large part to reconstructions using OUTBREAK [see papers cited by @Lynch2012], we now know considerably more about forest defoliator outbreak regimes than we did in the 1980s when OUTBREAK was under development. We think that two or more prolonged events separated by a single year should in some situations be considered a single event. This is particularly relevant to WSBW and spruce budworm (*C. fumiferana*), for which outbreak regimes have been reconstructed for several geographic areas and for which researchers and forest health experts have gained considerable knowledge [@Schmitt1984; @Sanders1985; @Brookes1987, and many later publications].  In many cases, the greatest growth suppression often occurs late in the outbreak due to cumulative effects accrued on a tree's resources [@Swetnam1989; @Swetnam1993]. We urge caution in using the bridging option, however, because it may not be appropriate for all studied insects, such as in situations where impacted stands barely recover from one outbreak before another begins, as with pine processionary caterpillars (*Thaumetopoea pityocampa* (Lepidoptera: Thaumetopoeidae)) [@Carus2004; @Carus2009].

(ref:BridgeCap) Examples of sequential defoliation events (bars), wherein two or three events occur separated by single years. In each case OUTBREAK recorded only one of the events, omitting those with a higher "max_defol" year, or the year of the lowest NGSI value. `dfoliatR`, by contrast, recorded all of the events, including two sequential events on DMJ26 and EFK22, and three on FNL04. Events recorded by both programs have a darker color, while those recorded only by `dfoliatR` are lighter in color. Adding the bridging option in `dfoliatR` links these events ("bridge_defol") into single longer events that may more accurately capture the effects of long-term insect herbivory on individual trees.

```{r fig-bridge, echo=FALSE, fig.cap='(ref:BridgeCap)'}
include_graphics(here("paper_elsevier", "Output", "bridging.pdf"))
```

@@

* *Sequential events* (36%) in which OUTBREAK omitted back-to-back events that occur one year prior or one year following an identified event. When this occurs, OUTBREAK selects the one event sequence with the lowest negative index year (e.g., Figure \@ref(fig:fig-bridge)). On two trees, OUTBREAK omitted two of three sequential events. While inspecting these differences, we added an option to `defoliate_trees()` that would "bridge" between sequential events (that surpass magnitude and duration thresholds) into single, long events. We felt that this was ecologically justified, especially for studies of WSBW, because the insect will occasionally defoliate individual trees for decades and reconstructed chronologies have show outbreaks as long as **xxx (ANN CHECK)**.

@@

At the site level, OUTBREAK and `dfoliatR` produce similar time series of percent trees defoliated (Figure \@ref(fig:fig-df-obr)), which forms the basis for inferring outbreak occurrence, intensity, and duration. In nearly site comparison, `dfoliatR` included either more events or longer durations of inferred outbreak. These differences arise from the inclusion of tree-level events by `dfoliatR` that were omitted by OUTBREAK. The differences translate to greater numbers of trees in defoliation during certain events, or events recorded by single trees during periods of low sample depth.

(ref:df-obrCap) Comparison of reconstructed western spruce budworm outbreaks computed by `dfoliatR` and OUTBREAK. Input parameters were identical between programs. Difference arise because `dfoliatR` will identify and record more defoliation events on individual trees, and may more accurately capture herbivory dynamics.

``` {r fig-df-obr, echo=FALSE, fig.cap='(ref:df-obrCap)'}
include_graphics(here("paper_elsevier", "Output", "plot-sites-df-obr.pdf"))
```

@@

The `dfoliatR` package provides dendroecologists with tools to infer, quantify, analyze, and visualize tree-ring growth suppression events and to reconstruct forest insect defoliator outbreak regimes.  It is built on the long-accepted host to non-host comparison methodology used in the 1980s FORTRAN program OUTBREAK [@Swetnam1985; @Swetnam1989; @Swetnam1993]. Our evaluation of the two programs revealed that `dfoliatR` excelled in identifying defoliation events on single trees, providing researchers with more consistent and biologically-justifiable results. Other key benefits are that `dfoliatR` provides easier control of suppression thresholds, additional output tables, high-quality and customizable graphics, and operates in the open-source `R` environment that is stable across computing platforms and under active development and maintenance by a large and growing community.

@@

<!-- <div id="refs"></div> -->
<!-- \newpage -->

<!-- # Tables & Figures {-} -->
